Classical Planning Algorithms ¡ on the Atari Video Games: Preliminary Results

### 2014-10-31 workshop-paper/paper.tex
Hector updates

%% Macros
\newcommand{\tuple}[1]{{\langle #1\rangle}}
\newcommand{\triple}[1]{{\langle #1\rangle}}
\newcommand{\pair}[1]{{\langle #1\rangle}}

@@

\newcommand{\eqdef}{\stackrel{\hbox{\tiny{def}}}{=}}
\newcommand{\IW}{{\textit{IW}}}
\newcommand{\SIW}{{\textit{SIW}}}
\newcommand{\ID}{{\textit{ID}}}
\newcommand{\BRFS}{{\textit{BrFS}}}
\newcommand{\BFS}{{\textit{2BFS}}}

@@

\begin{abstract}
\begin{quote}
The Atari  2600 games supported in  the Arcade Learning Environment (Bellemare et. al. 2013) 
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however, 
cannot be used  for selecting actions  for  two reasons: first,  no  compact PDDL-model of  the games is given, 
and  more importantly, the actions effects and the goals are  not known a priori. Moreover, in these games there is usually  
no set of goals to be achieved but rewards to be collected. These features do not preclude the use of   classical  algorithms like
breadth-first search or Dijkstra's algorithm,  but    these  algorithms are not effective over  large state spaces. We thus turn to a
different class of classical planning algorithms  introduced recently  that perform a \emph{structured exploration} of the state space; namely,  
like breadth-first search and Dijkstra's  algorithm they are ``blind'' and hence  do not require prior knowledge of state transitions, 
costs (rewards) or goals, and yet, like  heuristic search algorithms, they have been shown to be   effective for solving   problems over huge 
state spaces. The simplest such algorithm, called Iterated Width or IW, consists of  a sequence of calls  IW(1), IW(2), .., IW(k) where  IW(i) is  
a breadth-first search in which a state is  pruned when it is not the first state in the search  to make true some set of i atoms or less.
Preliminary empirical results suggest that the performance of IW with the  $k$ parameter fixed to $1$, i.e., IW(1), 
is much better than random or breadth-first search, and close to the performance of UCT. Other simple variations of IW
that combine exploration and exploitation are evaluated as well.
\end{quote}
\end{abstract}

@@

While all  these games feature a known initial (RAM) state and actions that
have deterministic effects on this state, the problem of selecting the next action
to be done cannot be addressed with state-of-the-art classical planners for two 
reasons. First, there is no compact PDDL-like encoding of the domain, and more 
importantly,  the goal to be achieved in each game  is not given. Indeed, there are 
often no goals at all but rewards $r(a,s)$ that potentially depend on the action $a$ and 
the state $s$ in  which  the action is performed, and these rewards are not known a priori either. 
Thus no model of the  goals  or the  rewards can be used to bias the search, and hence, 
standard heuristic search and classical planning methods can't be used.

The action selection problem in the Atari games can be seen as a \emph{reinforcement learning} problem 
\cite{sutton:book} over a deterministic MDP where the state transitions and rewards are not known, or alternatively,
as a  \emph{net-benefit planning problem} \cite{ref-net-benefit-planning} with unknown state transitions  and  rewards. Namely, 
we seek  an action sequence $a_0, a_1, \ldots, a_m$ from the given current  state $s_0$ that generates 
a state sequence $s_0, s_1, \ldots, s_{m+1}$ with  maximum  total reward $\sum_{i=0}^{m} r(a_i,s_i)$,  
where $m$ is a sufficiently large planning horizon, and the rewards $r(a_i,s_i)$ and the  state-transitions $s_{i+1}=f(a_i,s_i)$ 
are known only after the action $a_i$ is applied in the state $s_i$.

@@

The limitation of the   basic blind search  methods is that they are not effective  over very spaces, neither
for solving problems off-line, nor for guiding  a lookahead search for solving problems on-line. 
In this work, we thus turn to a recent class of  planning algorithms that combine the scope of blind search
methods with the performance of state-of-the-art classical planning algorithms: namely, like ``blind'' search algorithms
they do not require prior knowledge of state transitions, costs, or goals, and yet like heuristic algorithms
they manage to search large  state spaces effectively. The basic algorithm in this class is 
called IW for Iterated Width search \cite{nir:ecai2012}. IW consists of a sequence of calls  IW(1), IW(2), .., IW(k), where 
IW(i) is  a  standard breadth-first search where states are pruned right away when they fail to make true some new tuple (set) 
of at most $i$ atoms. Namely,   IW(1) is a breadth-first search that keeps  a state only if  the state is   the first  one 
in the search to make some atom  true; IW(2) keeps a state  only if the state is  the first one to  make  a pair of atoms true, and 
so on.  Like plain breadth-first  and iterative deepening searches,  IW  is 
complete, while searching the state space in a way that makes use of the 
\emph{structure of states} given by the values of a finite set of state variables.
In the Atari games, the (RAM) state is given by a vector of 128 bytes, which we associate with 128 variables 
$X_i$, $i=1, \ldots, 128$, each of which may take up to $256$ values $x_j$. A state $s$ makes an atom $X_i=x_{j}$ 
true when the value of the $i$-th byte in the state vector $s$ is $x_j$.

While a normal, complete breadth-first search runs in time that is exponential in the total number of variables $X_i$, 
the procedure $IW(k)$ runs in time that is exponential in the $k$ parameter, $1 \leq k \leq n$. 
Elsewhere, it has been shown that  IW exhibits   state-of-the-art performance on most existing planning benchmark domains 
when goals are restricted to be single atoms  where IW provably runs in low polynomial time in the number of
variables \cite{nir:ecai2012}. For example, in a a block world instance with any number of blocks and any initial
configuration,  the goal of  placing a given  block on top of another, will be achievable in  quadratic time  
in the number of blocks as any such instance can be shown to  have \emph{width} 2. This means  that IW(2) will be 
not only complete then, but also optimal. This is also true for most other benchmark domains in classical planning
that turn out to be have small, bounded widths for any instance, as long as the goal is atomic. 
For dealing with the benchmarks where goals are  not single atoms, e.g., where a given \emph{tower} of blocks needs to be built, 
simple extensions of the basic IW procedure have been developed as well, such as Serialized IW, where IW is 
used to achieve the goals one at at time \cite{nir:ecai2012,nir:ecai2014}. The extension of IW for problems where there is no goal but an
additive reward measure   to be optimized is direct.

@@

The sequence of calls IW(1), IW(2), \ldots, IW(k) manages to combine the scope of ``blind'' search algorithms with the performance of 
heuristic-search algorithms by exploring states  in an  order dictated by the ``width''  parameter $k$; 
states that have width 1 are explored first in linear time, then states that have  width two are explored in quadratic time, and so on. 
In classical planning problems, if the goal has a small width (something common in the classical planning domains when the  goal is a single atom, as
we have seen above), IW will find goal states  in low polynomial time. Similarly, in the reward-based setting, if  there are states of high reward 
that have small width, IW will find such states in low polynomial time, even if the state space is exponential in the number of variables. 
In this paper, we aim to test whether these ideas work for the Atari games too  when  the RAM state of the emulator
is used  as the system state.

The paper is organized as follows. We review the Iterated Width  algorithm and where it comes from, 
look at  some simple   variations of the basic algorithm that appear to be convenient  for the Atari
games, present and analyze the preliminary experimental results, summarize the main lessons, and
discuss related and future work.

@@

The Iterative Width  (IW) algorithm has been introduced as a classical planning algorithm
that takes a planning problem as an input, and computes an action sequence that solves the problem 
as the output \cite{nir:ecai2012}. The algorithm however
applies to a broader range of problems. We will characterize such
problems  by means  a finite and discrete set of states (the state space)
that correspond to vectors of size $n$. Namely, the 
states are  \emph{structured} or \emph{factored} , and we take each of the locations in the vector
to represent a variable $X_i$, and the value at that vector location to 
represent the value $x_j$ of variable $X_i$ in the state. 
In addition to the states space, a problem is defined by an initial state $s_0$, 
a set of actions $A(s)$ applicable in each state, a 
transition function $f$ such that  $s' = f(a,s)$ is 
the state that results from applying action $a \in A(s)$ to the state $s$, 
and rewards $r(a,s)$ represented by real numbers that result from  applying action $a$ in state $s$.
The transition and reward functions do not  need to be known a priori, yet in that case, the  
state and  reward that results from the application of an action in a state needs to be \emph{observable}.
The task is to compute an action sequence $a_0, \ldots, a_m$ for a large horizon $m$
that generates a state sequence $s_0, \ldots, s_{m+1}$ that  maximizes the accumulated 
reward $\sum_{i=0}^{m} r(a_i,s_i)$, or that provides a good approximation.

@@

\begin{itemize}
\item The \emph{novelty of a newly generate state $s$  in a  breadth-first search}
is   $1$ when $s$ is the first state generated in the search that makes true an atom $X=x$,
$2$  when $s$ is the first state that makes a pair of atoms $X=x$ and $Y=y$ true, and so on.

@@

For classical planning, the termination condition is the achievement of the goal. In the on-line setting, 
as in the Atari games, the termination condition is given by a time window or a maximum number of generated nodes.
The  best path found by IW is the path  that has a  maximum accumulated reward. 
The accummulated reward $R(s)$  of a state $s$ reached in an iteration of IW is determined by the  unique parent state $s'$ 
and action $a$ leading to $s$ from $s'$ as $R(s) = R(s') + r(a,s')$. The best state is the state $s$ with maximum reward
$R(s)$ generated but not pruned by IW, and the best path is the one that leads to the state $s$ from the current state.
The action selected to be done next is then the first action along such a path  (** see eventually randomization **).

@@

\caption{\small 
\emph{Top:} Number of classical planning problems  and percentages solved by IW(1), IW(2), or neither. Instances obtained
by splitting benchmarks  with $N$ atomic goals into $N$ problems with atomic goals.
\emph{Bottom:} Number of instances solved by \IW\ in comparison with  Iterative Deepening (\ID),  
Breadth-First Search (\BRFS),   and Greedy Best First Search (\textit{GBFS}) guided by  additive heuristic.
Time and  memory outs  after 30 minutes and 2 GB resp. From \cite{nir:ecai2012}.
}
\label{table1}
\end{table}
%%%%
%%%

@@

IW is a systematic and complete blind-search algorithm like breadth-first search (BrFS) and iterative deepening (ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the 
search in a different way. This structured exploration  has proved to be very effective in classical planning 
over most of the benchmark domains  when the goals are restricted to be single atoms. Table~\ref{table1} from \cite{nir:ecai2012} shows 
the percentage of instances that are  solved by  the first iteration of IW, i.e.  IW(1), by the second iteration 
IW(2), and by neither one.  These are instances that  have been obtained from the existing benchmarks
by splitting problems with $N$ atomic goals, into $N$ problems with one atomic goal each. As the table shows, 
$37\%$ of the 37921 instances are solved by IW(1) while $51.3\%$ are  solved by IW(2). Since IW($k$)
runs in time that is exponential in $k$, this mean that almost $90\%$ of the 37921 instances are  solved in time that
is either linear or quadratic in the number of problem variables, which in these encodings are all \emph{boolean}.  
Furthermore, when the performance of \IW\   is compared with  breadth-first search and iterative deepening, 
on the one hand,  and  with a  Greedy Best First Search guided by the additive heuritstic $h_{add}$
on the other (this algorithm is similar to one used by  the FF planner when hill climbing search fails \cite{hoffmann:ff}), 
it turns out that ``blind'' \IW\ solves as many problems as the informed search, 34627 vs. 34849, far ahead
of the other blind algorithms BrFS and ID  that solve 9010 and 8762 problems each. This is shown in the bottom
part of the Table~\ref{table1}. Moreover, \IW\ is  faster and results in shorter plans than the heuristic search  
\cite{nir:ecai2012}.

The min $k$ value for which $IW(k)$ solves an instance  is clearly bounded and small in most of
these instances. This is  actually no  accident and  has a \emph{theoretical explanation.} Lipovetzky and Geffner define
a structural parameter called the problem \emph{width} and show that for many of these domains,
\emph{any} solvable instance with atomic goals will  have a bounded and small width that is independent of the number
of variables and states in the problem. The min value $k$ for which the iteration  $IW(k)$ solves
the problem cannot exceed the problem width, so that the algorithm \emph{IW will run in time and space
that are exponential in the problem width.}

@@

While this notion of width and the iterated  width algorithms that are based on it  have been designed
for problems where a goal state needs to be reached, the notions remain  relevant  in  optimization problems as well.
Indeed, if a  good path is made of states $s_i$ each of which have a low width, we would like 
IW to find such path in low polynomial time for a low value of the $k$ parameter. 
Later on we will discuss the changes required in IW to enforce this property, which
are not yet implemented in the code that is evaluated below.

\section{Width-Based Algorithms for the Atari Games}

%%Explain: IW(1), 2BFS, and possible IW11. Also, Dijkstra postprocessing over IW(1) graph if ran.

The number of nodes \emph{generated} by IW(1) is $n \times D \times b$ in the worst case, 
where $n$ is the number of variables, $D$ is the size of their domains, and $b$ is maximum the number of 
actions applicable per  state.  This same number in breadth-first search is not linear in $n$
but exponential. For the Atari games,  $n=128$, $D=256$, and $b=18$, so that the product is
equal to $589,824$, which is large but feasible. On the other hand, the number
of nodes generated by IW(2) in the worst case is  $(n \times D)^2 \times b$, which is
equal to $1,073,741,824$ which is too large, forcing us to consider only a fraction of 
such states. For classical planning problems, the growth in the number of nodes from 
IW(1) to IW(2) is not that large, as the variables are boolean. Indeed, we could have
taken the state vector for the Atari games as a vector of 1024 boolean variables, 
and apply these algorithms to that representation. The results would be different.
In such a case, IW(1) would generate $n' \times D' \times b$ states, and 
IW(2), $(n' \times D')^2 \times b$ states, which for $n'=1024$ and $D=2$
represent  $36,864$ and $7,5497,472$ states respectively.
These are both feasible numbers, but we haven't tried yet this representation,
under the assumption that the correlations among the bits in each one of 
the 128 words of the state vectors are meaningful. 
In summary, from the basic IW algorithm we are testing only the first 
call, IW(1).

IW is a purely exploration algorithm that does not take into account  the accumulated 
reward for selecting the next states to consider. As a simple, variant that combines
exploration and exploitation, we evaluated a \emph{best-first search} algorithm
with two queues: one queue  ordered first by  novelty measure 
(recall, novelty 1 means that the state is the first one to make some atom true),
and a second queue ordered by accumulated reward. In one iteration, the best first
search picks up the best node from one queue, and in the second iteration it picks
up the best node from the other queue. This way for combining multiple heuristics
is used in the LAMA planner \cite{richter:lama}, and was introduced in the
planner Fast Downward \cite{helmert:fd,malte:alternating}. In addition, we break ties
in the first queue favoring states with largest accumulated reward, and 
in the second queue, favoring states with smallest novelty measure. Last, when
a node is expanded, it is removed from the queue, and its children are placed on 
both queues. The exception are the nodes with accumulated reward 0 that are only
placed in the first queue.
We refer to this best-first algorithm below, as 2BFS.

In the experiments below, we use IW(1) with two simple variations.
First, in the breadth-first search, we generate the children in 
random order. This makes the execution that results from the
IW(1) lookahead less susceptible to fall into loops; a potential
problem in local search algorithms with no memory or learning.
Second, a discount factor $\gamma=0.995$ is added
for discounting future rewards like in UCT. For this,
each state $s$ keeps its depth d(s) in the IW(1) tree,
and if state s' is the child of state s and action a, 
$R(s')$ is set to $R(s) + \gamma^d(s´) r(a,s)$.
A discount factor less than 1 results in a slight
preference for rewards that can be reached earlier, 
which is reasonable  in the on-line setting.

%% NIR: IW11 Don't need to be explained. Maybe in Future Work?

\section{Experimental Results in the Atari Games}

\input{performance_table}

\input{experimental_results}

\section{Exploration and Exploitation: Discussion}

The notion of width underlying the iterated width algorithm was developed in the context
of classical planning in order to understand why most of the hundreds of existing benchmarks
resulting from planning competitions appear to be relatively simple for current planners, even
though classical planning is PSPACE-complete \cite{bylander:complexity}. The partial answer is that 
most of these domains have a low width, and hence, can be solved in low polynomial time (by IW),
when goals contain a single atom. At the same time, most problems with multiple atomic goals are 
also easy too, as the goals can often be achieved one at a time \cite{nir:ecai2012}.

In the Iterated Width algorithm, the key notion is the \emph{novelty measure} of a state in the underlying
breadth-first search. In particular, a state gets a novelty 1 measure in the search if it makes true,
some atom that up to that point had been false in all previously generated states. In the procedure IW(1)
all other states are pruned right away, as if they were ``duplicate'' states. The algorithms IW and IW(k)
are pure  exploration algorithms that make no use at all of the information about the goal, if there is a goal,
or about the  reward function to optimize otherwise. They are  thus  pure exploration algorithms guided by the novelty
measures, which use the structure of the states to provide a  structure to the exploration of the state space,
which in the case of IW is guaranteed to be complete.

>>>>

\newcommand{\tuple}[1]{{\langle #1\rangle}}
\newcommand{\triple}[1]{{\langle #1\rangle}}
\newcommand{\pair}[1]{{\langle #1\rangle}}

@@

\newcommand{\propername}[1]{\textsc{#1}}
\newcommand{\eqdef}{\stackrel{\hbox{\tiny{def}}}{=}}
%\newcommand{\IW}{{\textit{IW}}}
%\newcommand{\SIW}{{\textit{SIW}}}
%\newcommand{\ID}{{\textit{ID}}}
%\newcommand{\BRFS}{{\textit{BrFS}}}
%\newcommand{\BFS}{{\textit{2BFS}}}
%\newcommand{\UCT}{{\textit{UCT}}}
\newcommand{\IW}{{\textsc{IW}}}
\newcommand{\SIW}{{\textsct{SIW}}}
\newcommand{\ID}{{\textsc{ID}}}
\newcommand{\BRFS}{{\textsc{BrFS}}}
\newcommand{\BFS}{{\textsc{2BFS}}}
\newcommand{\UCT}{{\textsc{UCT}}}

\newcommand{\Asterix}{\propername{Asterix}}
\newcommand{\BeamRider}{\propername{Beam Rider}}
\newcommand{\Breakout}{\propername{Breakout}}
\newcommand{\Enduro}{\propername{Enduro}}
\newcommand{\Freeway}{\propername{Freeway}}
\newcommand{\Pong}{\propername{Pong}}
\newcommand{\Qbert}{\propername{Q*Bert}}
\newcommand{\Seaquest}{\propername{Seaquest}}
\newcommand{\SpaceInvaders}{\propername{Space Invaders}}
\newcommand{\Deepmind}{\propername{Deepmind}}

@@

\begin{abstract}
\begin{quote}
The Atari  2600 games supported in  the Arcade Learning Environment (Bellemare et. al. 2013) 
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however, 
cannot be used  for selecting actions  for  two reasons: first,  no  compact PDDL-model of  the games is given, 
and  more importantly,  the action effects and goals are  not known a priori. Moreover, in these games there is usually  
no set of goals to be achieved but rewards to be collected. These features do not preclude the use of   classical  algorithms like
breadth-first search or Dijkstra's algorithm,  but    these methods are not effective over  large state spaces. We thus turn to a
different class of classical planning algorithms  introduced recently  that perform a \emph{structured exploration} of the state space; namely,  
like breadth-first search and Dijkstra's  algorithm they are ``blind'' and hence  do not require prior knowledge of state transitions, 
costs (rewards) or goals, and yet, like  heuristic search algorithms, they have been shown to be   effective for solving   problems over huge 
state spaces. The simplest such algorithm, called Iterated Width or IW, consists of  a sequence of calls  IW(1), IW(2), .., IW(k) where  IW(i) is  
a breadth-first search in which a state is  pruned when it is not the first state in the search  to make true some set of i atoms or less.
Preliminary empirical results suggest that the performance of IW with the  $k$ parameter fixed to $1$, 
i.e., IW(1), is much better than random or breadth-first search, and approaches 
the performance of UCT. A simple best-first variation  of IW,  that combine exploration and exploitation,
is evaluated as well. 
\end{quote}
\end{abstract}

@@

While all  these games feature a known initial (RAM) state and actions that
have deterministic effects on the state, the problem of selecting the next action
to be done cannot be addressed with state-of-the-art classical planners \cite{geffner:2013book}.
This is because there is  no compact PDDL-like encoding of the domain, and more 
importantly,  the goal to be achieved in each game  is not given. Indeed, there are 
often no goals  but rewards $r(a,s)$ that  depend on the action $a$ and 
the state $s$ in  which  the action is performed, and these rewards are not known a priori 
either.\footnote{Actually, in the Atari games, the rewards $r(a,s)$ depend only
on the state $f(a,s)$ that results from doing action $a$ in the state $s$.}
Thus no model of the  goals  or the  rewards can be used to bias the search, and hence, 
standard heuristic search and classical planning methods can't be used.

The action selection problem in the Atari games can be seen as a \emph{reinforcement learning} problem 
\cite{sutton:book} over a deterministic MDP where the state transitions and rewards are not known, or alternatively,
as a  \emph{net-benefit planning problem} \cite{ref-net-benefit-planning,keyder} with unknown state transitions  and  rewards. Namely, 
we seek  an action sequence $a_0, a_1, \ldots, a_m$ from the given current  state $s_0$ that generates 
a state sequence $s_0, s_1, \ldots, s_{m+1}$ with  maximum  total reward $\sum_{i=0}^{m} r(a_i,s_i)$,  
where $m$ is a sufficiently large planning horizon, and the rewards $r(a_i,s_i)$ and the  state-transitions $s_{i+1}=f(a_i,s_i)$ 
are known only after  action $a_i$ is applied in the state $s_i$.

@@

The limitation of the   basic blind search  methods is that they are not effective  over large  spaces, neither
for solving problems off-line, nor for guiding  a lookahead search for solving problems on-line. 
In this work, we thus turn to a recent class of  planning algorithms that combine the scope of blind search
methods with the performance of state-of-the-art classical planning algorithms: namely, like ``blind'' search algorithms
they do not require prior knowledge of state transitions, costs, or goals, and yet like heuristic algorithms
they manage to search large  state spaces effectively. The basic algorithm in this class is 
called IW for Iterated Width search \cite{nir:ecai2012}. IW consists of a sequence of calls  IW(1), IW(2), .., IW(k), where 
IW(i) is  a  standard breadth-first search where states are pruned right away when they fail to make true some new tuple (set) 
of at most $i$ atoms. Namely,   IW(1) is a breadth-first search that keeps  a state only if  the state is   the first  one 
in the search to make some atom  true; IW(2) keeps a state  only if the state is  the first one to  make  a pair of atoms true, and 
so on.  Like plain breadth-first  and iterative deepening searches,  IW  is 
complete, while searching the state space in a way that makes use of the 
\emph{structure of states} given by the values of a finite set of state variables.
In the Atari games, the (RAM) state is given by a vector of 128 bytes, which we associate with 128 variables 
$X_i$, $i=1, \ldots, 128$, each of which may take up to $256$ values $x_j$. A state $s$ makes an atom $X_i=x_{j}$ 
true when the value of the $i$-th byte in the state vector $s$ is $x_j$.

While a normal, complete breadth-first search runs in time that is exponential in the total number of variables $X_i$, 
the procedure $IW(k)$ runs in time that is exponential in the $k$ parameter, $1 \leq k \leq n$. 
Elsewhere, it has been shown that  IW exhibits   state-of-the-art performance on most existing planning benchmark domains 
when goals are restricted to be single atoms  where IW provably runs in low polynomial time in the number of
variables \cite{nir:ecai2012}. For example, in a block world instance with any number of blocks and any initial
configuration,  the goal of  placing a given  block on top of another will be achievable in  quadratic time  
in the number of blocks as any such instance can be shown to  have \emph{width} 2. This means  that IW(2) will be 
not only complete then, but also optimal. This is also true for most other benchmark domains in classical planning
that turn out to  have small, bounded widths for any instance, as long as the goal is atomic. 
For dealing with the benchmarks where goals are  not single atoms, e.g., where a given \emph{tower} of blocks needs to be built, 
simple extensions of the basic IW procedure have been developed as well, such as Serialized IW, where IW is 
used to achieve the goals one at at time \cite{nir:ecai2012,nir:ecai2014}. 
The extension of IW for problems where there is no goal but an
additive reward measure  to be optimized, like in the Atari games, is direct.

@@

\Omit{
The sequence of calls IW(1), IW(2), \ldots, IW(k) manages to combine the scope of ``blind'' search algorithms with the performance of 
heuristic-search algorithms by exploring states  in an  order dictated by the ``width''  parameter $k$; 
states that have width 1 are explored first in linear time, then states that have  width two are explored in quadratic time, and so on. 
In classical planning problems, if the goal has a small width (something common in the classical planning domains when the  goal is a single atom, as
we have seen above), IW will find goal states  in low polynomial time. Similarly, in the reward-based setting, if  there are states of high reward 
that have small width, IW will find such states in low polynomial time, even if the state space is exponential in the number of variables. 
In this paper, we aim to test whether these ideas work for the Atari games too  when  the RAM state of the emulator
is used  as the system state. 
}

The paper is organized as follows. We review the iterated width  algorithm and where it comes from, 
look at  simple   variations of the algorithm that appear to be convenient  for the Atari
games, present and analyze the preliminary experimental results, and discuss related and future work.

@@

The Iterative Width  (IW) algorithm has been introduced as a classical planning algorithm
that takes a planning problem as an input, and computes an action sequence that solves the problem 
as the output \cite{nir:ecai2012}. The algorithm however
applies to a broader range of problems. We will characterize such
problems  by means  a finite and discrete set of states (the state space)
that correspond to vectors of size $n$. Namely, the 
states are  \emph{structured} or \emph{factored} , and we take each of the locations in the vector
to represent a variable $X_i$, and the value at that vector location to 
represent the value $x_j$ of variable $X_i$ in the state. 
In addition to the state space, a problem is defined by an initial state $s_0$, 
a set of actions applicable in each state, a 
transition function $f$ such that  $s' = f(a,s)$ is 
the state that results from applying action $a \in A(s)$ to the state $s$, 
and rewards $r(a,s)$ represented by real numbers that result from  applying action $a$ in state $s$.
The transition and reward functions do not  need to be known a priori, yet in that case, the  
state and  reward that results from the application of an action in a state needs to be \emph{observable}.
The task is to compute an action sequence $a_0, \ldots, a_m$ for a large horizon $m$
that generates a state sequence $s_0, \ldots, s_{m+1}$ that  maximizes the accumulated 
reward $\sum_{i=0}^{m} r(a_i,s_i)$, or that provides a good approximation.

@@

\begin{itemize}
\item The \emph{novelty of a newly generate state $s$  in a  breadth-first search}
is   $1$ if  $s$ is the first state generated in the search that makes true some  atom $X=x$, 
else it is $2$ if  $s$ is the first state that makes a \emph{pair} of atoms $X=x$ and $Y=y$ true, 
and so on.

@@

For classical planning, the termination condition is the achievement of the goal. In the on-line setting, 
as in the Atari games, the termination condition is given by a time window or a maximum number of generated nodes.
The  best path found by IW is the path  that has a  maximum accumulated reward. 
The accummulated reward $R(s)$  of a state $s$ reached in an iteration of IW is determined by the  unique parent state $s'$ 
and action $a$ leading to $s$ from $s'$ as $R(s) = R(s') + r(a,s')$. The best state is the state $s$ with maximum reward
$R(s)$ generated but not pruned by IW, and the best path is the one that leads to the state $s$ from the current state.
The action selected to be done next is then the first action along such a path.

@@

\caption{\small 
\emph{Top:} Number of classical planning problems  and percentages solved by IW(1), IW(2), or neither. Instances obtained
by splitting benchmarks  with $N$ atomic goals into $N$ problems with atomic goals.
\emph{Bottom:} Number of instances solved by \IW\ in comparison with  Iterative Deepening (\ID),  
Breadth-First Search (\BRFS),   and Greedy Best First Search (\textit{GBFS}) guided by  additive heuristic.
Time and  memory outs  after 30 minutes and 2 GB resp. From \cite{nir:ecai2102}.
}
\label{table1}
\end{table}
%%%%
%%%

@@

IW is a systematic and complete blind-search algorithm like breadth-first search (BrFS) and iterative deepening (ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the 
search in a different way. This structured exploration  has proved to be very 
effective over classical planning benchmark domains  when goals are  single atoms.\footnote{
Any conjunctive goal can be mapped into a single dummy atomic goal by adding an action that
achieves the dummy goal and that has the original conjunctive goal as a precondition. 
Yet, this changes the domain.}
Table~\ref{table1} from \cite{nir:ecai2012} shows 
the percentage of instances that are  solved by  the first iteration of IW, i.e.  IW(1), by the second iteration 
IW(2), and by neither one.  These are instances that  have been obtained from the existing benchmarks
by splitting problems with $N$ atomic goals, into $N$ problems with one atomic goal each. As the table shows, 
$37\%$ of the 37921 instances are solved by IW(1) while $51.3\%$ are  solved by IW(2). Since IW($k$)
runs in time that is exponential in $k$, this mean that almost $90\%$ of the 37921 instances are  solved in time that
is either linear or quadratic in the number of problem variables, which in these encodings are all \emph{boolean}.  
Furthermore, when the performance of \IW\   is compared with  breadth-first search and iterative deepening, 
on the one hand,  and  with a  Greedy Best First Search guided by the additive heuristic $h_{add}$ 
\cite{bonet:hsp-aij} on the other (this algorithm is similar 
to one used by  the FF planner when hill climbing search fails \cite{hoffmann:ff}), 
it turns out that ``blind'' \IW\ solves as many problems as the informed search, 34627 vs. 34849, far ahead
of the other blind algorithms BrFS and ID  that solve 9010 and 8762 problems each. This is shown in the bottom
part of the Table~\ref{table1}. Moreover, \IW\ is  faster and results in shorter plans than the heuristic search  
\cite{nir:ecai2012}.

The min $k$ value for which $IW(k)$ solves a problem  is clearly bounded and small in most of
these instances. This is  actually no  accident and  has a \emph{theoretical explanation.} Lipovetzky and Geffner define
a structural parameter called the problem \emph{width} and show that for many of these domains,
\emph{any} solvable instance with atomic goals will  have a bounded and small width that is independent of the number
of variables and states in the problem. The min value $k$ for which the iteration  $IW(k)$ solves
the problem cannot exceed the problem width, so  the algorithm \emph{IW runs in time and space
that are exponential in the problem width.}

@@

While this notion of width and the iterated  width algorithms that are based on it  have been designed
for problems where a goal state needs to be reached, the notions remain  
relevant  in  optimization problems as well. Indeed, if a  good path is made of 
states $s_i$ each of which have a low width, IW can be made 
to find such path in low polynomial time for a small value of the $k$ parameter.
Later on we will discuss  a slight  change  required in IW to enforce this property
which is actually  not yet implemented in the code that is tested below.

\section{The Algorithms for the Atari Games}

The number of nodes \emph{generated} by IW(1) is $n \times D \times b$ in the worst case, 
where $n$ is the number of variables, $D$ is the size of their domains, and $b$ is the number of 
actions per  state.  This same number in breadth-first search is not linear in $n$
but exponential. For the Atari games,  $n=128$, $D=256$, and $b=18$, so that the product is
equal to $589,824$, which is large but feasible. On the other hand, the number
of nodes generated by IW(2) in the worst case is  $(n \times D)^2 \times b$, which is
equal to $1,073,741,824$ which is too large, forcing us to consider only a fraction of 
such states. For classical planning problems, the growth in the number of nodes from 
IW(1) to IW(2) is not that large, as the variables are boolean. Indeed, we could have
taken the state vector for the Atari games as a vector of 1024 boolean variables, 
and apply these algorithms to that representation. The results would be different.
In such a case, IW(1) and IW(2)  would generate $n' \times D' \times b$ and 
$(n' \times D')^2 \times b$ states, which for $n'=1024$ and $D=2$
represent  $36,864$ and $7,5497,472$ states respectively.
These are feasible numbers, but we haven't tried  this representation yet
under the assumption that the correlations among the bits in each one of 
the 128 words of the state vectors are meaningful. 
In summary, from the basic IW algorithm we are testing only the first 
call IW(1).

IW is a purely exploration algorithm that does not take into account  the accumulated 
reward for selecting the next states to consider. As a simple, variant that combines
exploration and exploitation, we evaluated a \emph{best-first search} algorithm
with two queues: one queue  ordered first by  novelty measure 
(recall that novelty one means that the state is the first one to make some atom true),
and a second queue ordered by accumulated reward. In one iteration, the best first
search picks up the best node from one queue, and in the second iteration it picks
up the best node from the other queue. This way for combining multiple heuristics
is used in the LAMA planner \cite{richter:lama}, and was introduced in the
planner Fast Downward \cite{helmert:fd,malte:alternating}. In addition, we break ties
in the first queue favoring states with largest accumulated reward, and 
in the second queue, favoring states with smallest novelty measure. Last, when
a node is expanded, it is removed from the queue, and its children are placed on 
both queues. The exception are the nodes with no accumulated reward  that are 
placed in the first queue only. We refer to this best-first algorithm 
as 2BFS.

For  the experiments below, we added two simple variations to  IW(1) and 2BFS.
First, in the breadth-first search underlying IW(1), we generate the children in 
random order. This makes the executions that results from the
IW(1) lookahead less susceptible to fall into loops; a potential
problem in local search algorithms with no memory or learning.
Second, a discount factor $\gamma=0.995$ is used in both algorithms 
for discounting future rewards like in UCT. 
For this, each state $s$ keeps its depth $d(s)$ in the search tree,
and if state s' is the child of state $s$ and action $a$, 
$R(s')$ is set to $R(s) + \gamma^d(s´) r(a,s)$.
The  discount factor  results in a slight
preference for rewards that can be reached earlier, 
which is a  reasonable heuristic in  on-line settings
based on lookahead searches.

\section{Experimental Results}
